---
title: "Multivariate demo"
output: 
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r}
library(tidyverse)        #for visualization and wrangling
library(ggridges)         #for making density ridge plots
library(palmerpenguins)   #for penguins data
```

Load the data:

```{r}
data("penguins")
```


## Quantitative vs. quantitative

Examine the relationship between `body_mass_g` and `flipper_length_mm` via a scatterplot.

```{r}

```

And, let's describe this relationship in terms of form, strength, direction, and calling out any outlying points. 

Computing the correlation can be a convenient way to numerically describe the strength and direction of the LINEAR relationship.

```{r}

```

## Quantitative vs. categorical

Now, we want to examine the relationship between `body_mass_g` and `island`. We can phrase this as examining how the distribution of `body_mass_g` varies by `island`. We'll do this first with a side-by-side boxplot (I'll explain what is show in the boxplot).

```{r}

```

We can also do this using a density ridge plot. This uses `geom_density_ridges()` from the `ggridges` library.

```{r}

```

For both of these, we want to describe what we're seeing. It can be helpful to compute some statistics for each level of the categorical variable to help us. 

```{r}

```







